---
title: "Exploratory Data Analysis: Socio-Economic Predictors of Malaria"
author: "Applied Statistics Portfolio Project"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 2
---

## Introduction

This report performs exploratory data analysis (EDA) for a malaria epidemiology dataset, with a focus on socio-economic predictors. The objective is to characterize variable distributions, assess data quality, and visualize relationships that motivate multivariable logistic regression.

```{r setup, message=FALSE, warning=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.width = 9, fig.height = 6, dpi = 300)
```

## Methodology

### Load Packages and Data

```{r load-packages, message=FALSE}
required_packages <- c(
  "tidyverse",
  "janitor",
  "GGally",
  "scales"
)

missing_packages <- required_packages[!(required_packages %in% installed.packages()[, "Package"])]
if (length(missing_packages) > 0) {
  install.packages(missing_packages, dependencies = TRUE)
}

invisible(lapply(required_packages, library, character.only = TRUE))
```

```{r load-data}
data_path <- file.path("data", "malaria_data.csv")
output_dir <- "outputs"
if (!dir.exists(output_dir)) {
  dir.create(output_dir, recursive = TRUE)
}

malaria_data <- readr::read_csv(data_path, show_col_types = FALSE) |>
  janitor::clean_names()

numeric_vars <- malaria_data |>
  dplyr::select(where(is.numeric))
```

### Key Data Objects (Printed Output)

```{r object-output-data-preview}
dim(malaria_data)
names(malaria_data)
head(malaria_data, 10)
```

```{r object-output-summary}
summary(malaria_data)
```

```{r object-output-missingness}
missingness_table <- malaria_data |>
  summarise(across(everything(), ~ sum(is.na(.)))) |>
  tidyr::pivot_longer(cols = everything(), names_to = "variable", values_to = "missing_count") |>
  arrange(desc(missing_count))

missingness_table
```

```{r object-output-numeric-correlation}
if (ncol(numeric_vars) > 1) {
  cor_matrix <- stats::cor(numeric_vars, use = "pairwise.complete.obs")
  cor_matrix
} else {
  cor_matrix <- matrix(NA_real_, nrow = 1, ncol = 1, dimnames = list("insufficient_numeric_columns", "insufficient_numeric_columns"))
  cor_matrix
}
```

## Results

### Visualization 1: Distribution of Household Income by Malaria Status

```{r income-plot}
income_plot <- malaria_data |>
  dplyr::mutate(
    malaria_status = factor(
      malaria_status,
      levels = c(0, 1),
      labels = c("Negative", "Positive")
    )
  ) |>
  ggplot2::ggplot(ggplot2::aes(x = household_income, fill = malaria_status)) +
  ggplot2::geom_histogram(
    bins = 35,
    alpha = 0.7,
    position = "identity",
    color = "white"
  ) +
  ggplot2::scale_fill_brewer(palette = "Set1", name = "Malaria Status") +
  ggplot2::scale_x_continuous(labels = scales::label_comma()) +
  ggplot2::labs(
    title = "Distribution of Household Income by Malaria Status",
    x = "Household Income",
    y = "Count"
  ) +
  ggplot2::theme_minimal(base_size = 13) +
  ggplot2::theme(plot.title = ggplot2::element_text(face = "bold"))

income_plot
```

```{r save-income-plot, echo=TRUE}
ggplot2::ggsave(
  filename = file.path(output_dir, "eda_income_distribution.png"),
  plot = income_plot,
  width = 9,
  height = 6,
  dpi = 300
)
```

### Visualization 2: Correlation Heatmap of Numeric Variables

```{r correlation-heatmap}
cor_df <- as.data.frame(as.table(cor_matrix))
colnames(cor_df) <- c("Var1", "Var2", "Correlation")

correlation_plot <- ggplot2::ggplot(cor_df, ggplot2::aes(Var1, Var2, fill = Correlation)) +
  ggplot2::geom_tile(color = "white") +
  ggplot2::scale_fill_gradient2(
    low = "#2166AC",
    mid = "white",
    high = "#B2182B",
    midpoint = 0,
    limits = c(-1, 1),
    na.value = "grey90"
  ) +
  ggplot2::labs(
    title = "Correlation Heatmap of Epidemiological Variables",
    x = NULL,
    y = NULL
  ) +
  ggplot2::theme_minimal(base_size = 12) +
  ggplot2::theme(
    axis.text.x = ggplot2::element_text(angle = 45, hjust = 1),
    plot.title = ggplot2::element_text(face = "bold")
  )

correlation_plot
```

```{r save-correlation-plot}
ggplot2::ggsave(
  filename = file.path(output_dir, "eda_correlation_heatmap.png"),
  plot = correlation_plot,
  width = 9,
  height = 7,
  dpi = 300
)
```

### Visualization 3: Estimated Malaria Probability Across Education Years

```{r education-plot}
education_plot <- malaria_data |>
  dplyr::mutate(malaria_status = as.numeric(malaria_status)) |>
  ggplot2::ggplot(ggplot2::aes(x = education_years, y = malaria_status)) +
  ggplot2::geom_jitter(
    alpha = 0.15,
    width = 0.2,
    height = 0.04,
    color = "#666666"
  ) +
  ggplot2::geom_smooth(
    method = "glm",
    method.args = list(family = "binomial"),
    se = TRUE,
    color = "#1B7837",
    fill = "#A6DBA0",
    linewidth = 1.1
  ) +
  ggplot2::scale_y_continuous(
    breaks = c(0, 0.25, 0.5, 0.75, 1),
    labels = scales::label_percent(accuracy = 1)
  ) +
  ggplot2::labs(
    title = "Estimated Malaria Probability Across Education Years",
    x = "Years of Education",
    y = "Predicted Malaria Probability"
  ) +
  ggplot2::theme_minimal(base_size = 13) +
  ggplot2::theme(plot.title = ggplot2::element_text(face = "bold"))

education_plot
```

```{r save-education-plot}
ggplot2::ggsave(
  filename = file.path(output_dir, "eda_malaria_vs_education.png"),
  plot = education_plot,
  width = 9,
  height = 6,
  dpi = 300
)
```

### Export Quick EDA Summary

```{r quick-summary}
eda_quick_summary <- malaria_data |>
  dplyr::summarise(
    n = dplyr::n(),
    malaria_prevalence = mean(malaria_status, na.rm = TRUE)
  )

eda_quick_summary

readr::write_csv(
  eda_quick_summary,
  file.path(output_dir, "eda_quick_summary.csv")
)
```
